{\LARGE Basis Choice for Scalar-on-Function Regression \\ with Applications to Near-Infrared Spectroscopy}

### 2022-02-09 Paper/main.tex
Take out subsections from Outlook to reduce Contents Length
Adjusted Capitalization of Monomial basis

	Due to the implementation of our simulation, we will limit our paper to the case of $\alpha = 0$. However, a different choice of $\alpha$ could lead to improvements in performance of the monomial basis.
	As the monomials are not pairwise orthogonal, this basis is prone to collinearity problems, which can result in numerically unstable estimates. This restricts the number of basis functions that can be used in the estimation procedures limiting its ability to capture pronounced local peculiarities. These problems will be addressed in more detail in later parts of this paper. The limitation to a low number of monomials can additionally lead to undesirable behavior away from the point of evaluation (cf. \cite{ramsay_functional_2005}).
	\vspace{-0.2cm}

	\paragraph{Fourier Basis}
	In the same way the monomial basis is connected to the Taylor series, the Fourier basis corresponds to the Fourier series, which can be used to decompose a periodic function into trigonometric functions. Equation \ref{Fourier_Series} shows an example for a function with a period of 1.
	\begin{equation}\label{Fourier_Series}
		s(x) = \frac{A_0}{2} + \sum_{i = 1}^{\infty} A_i \cos(2\pi i x - \phi_i) = \frac{a_0}{2} + \sum_{i = 1}^{\infty}\left[a_i \cos(2\pi i x) + b_i \sin(2\pi i x)\right]
	\end{equation}

@@

	As the B-spline basis does not have infinitely many elements as the full monomial or Fourier bases do, it is slightly misleading to speak of truncating the B-spline basis at a truncation parameter $L$ in later parts of this paper. For the sake of keeping the notation reasonably concise, we will still keep this notation. However, by convention, truncating a B-spline basis at truncation parameter $L$ shall mean using a B-spline basis consisting of $L$ functions from this point on.

@@

		where $Z \sim \mathcal{N}(0,1)$. In total, we created four combinations, using the two true coefficient functions and the two error terms. These four combinations are then used with a different number of monomial basis functions $ \in \{1,2, \dots, 6\}$, cubic B-spline basis-function $\{5,6,...,18\}$ and Fourier functions $\{1,3,...,25\}$ to predict the generated responses using the basis expansion approach and the FPCR approach.
		% For the evaluation, we used the prediction MSE calculated by 10 fold cross-validation.
		To obtain valid out of sample properties for the FPCR, within each of the ten ten-fold cross-validation splits, we first calculate the first $n_{FPC}$ FPC's for $n_{FPC} \in \{2,3,4\}$ of the training set $\mathcal{T}$, which was prior smoothed with the respective basis function specification. The obtained Eigenfunctions $\nu^{m,\mathcal{T}}$ are then used to estimate the scores of the holdout set $\mathcal{H}$, $\hat{\xi}_{i}^{m, \mathcal{H}}$  by the equation
		\begin{equation}
			\begin{split}
				\hat{\xi}_{i}^{m, \mathcal{H},L} & =  \int_{0}^{1} {\color{red} \left(X_{i}^{\mathcal{H}}(s) - \hat{\mu}^{\mathcal{T}}(s)\right)} {\color{blue}\tilde{\nu}^{m, L, \mathcal{T}} }\mathrm{d}s) 
			    \approx \int_{0}^{1} {\color{red}\left(\sum_{j = 1}^{L} a_{i,j}^{\mathcal{H}} \phi_j(s)\right)} {\color{blue}\left(\sum_{k = 1}^{L} d_{k}^{m, \mathcal{T}} \phi_{k}(s)\right)} \mathrm{d}s \\
				 &\approx \sum_{j = 1}^{L} \left[ a_{i,j}^{\mathcal{H}}\sum_{k = 1}^{L}  d_{k}^{m, \mathcal{T}} \int_{0}^{1} \phi_j(s) \psi_{k}(s)\mathrm{d}s\right]
			\end{split}
		\end{equation}

@@

	\paragraph{Monomial Basis}
	Due to the in the number of monomial basis functions increasing collinearity, simulations were conducted up until the sixth monomial, which already shows signs of numerical instability. For reasons outlined in \ref{Monomial_basis}, they seem suited for $f_1$ where it shows a better performance than B-splines for this assumed true coefficient function. A hypothesis for this could be that $f_1$ is an entire function, which can be well approximated with a power series. In case of $f_2$, this basis shows the weakest performance out of all basis, for which the plots in figure \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2} provide evidence. It seems like $\hat{\beta}(t)$ is not changing in the amount of noise and shows exaggerated behaviour at the boundaries. This weakness is especially pronounced in the MSPE for $f_2,Y_1$. For $f_1$, the simulation selects 5(3) and for $f_2$ 5(5) monomial basis functions for the high(low) signal-to-noise ratio. 
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	Simulations with B-spline basis functions were possible from 4 to 18, since from 18 onwards the simulations were running into collinearity problems.
	Function $f_1$ requires 5(4) B-spline basis functions for the high(low) signal-to-noise ratio to obtain the best fit for the B-spline basis and shows the worst performance of the three basis. An explanation might be its overripe behavior at the boundaries and the exaggeration of the peculiarities of $f_1$ ( figure \ref{basis_expansion_1_1} and \ref{basis_expansion_1_2}), which is especially pronounced for the lower signal-to-noise ratio. For $f_2$, 11(6) B-spline basis functions are chosen for the high(low) signal-to-noise ratio. The B-spline basis outperforms the monomial basis in $f_2$, but comes second to the Fourier basis. While the basis seems to recognize the peculiarities in $f_2,Y_1$, it seems to struggle with it for the noisy responses in $f_2,Y_2$ ( figure \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2}). With the low signal-to-noise ratio in the responses, the simulation chooses a smaller number of basis functions, which could be to prevent fitting this scalar noise using $\hat{\beta}(t)$.
	\vspace{-0.2cm}

@@

	\paragraph{Monomial Basis}
	For $f_1$, the cross-validated MSPE is decreasing in the number of FPC's and chosen basis functions for both the high and the low signal-to-noise ratio ($4,5,6$ basis functions for $n_{FPC} \in \{2, 3, 4 \}$). In  $f_2$ we also observe a decreasing MSPE in the number of FPC's, but no clear relationship for the chosen basis functions. The monomial basis shows the weakest performance for all three bases in each setting for all $n_{FPC}$.
	\vspace{-0.2cm}

@@

	\subsection{Interpretation of Results}
		\subsubsection{Basis Expansion Regression}
		The cross-validated MSPE selects 5 basis functions for the monomial basis. For B-Splines, the cross-validation selects 10 basis functions and 9 basis functions for the Fourier basis. Comparing the estimates in Appendix \ref{Estimates_Appl}, away from the boundaries, the B-spline and Fourier bases show similar behavior, opposing to the monomial estimate which might be attributable to the lower number of basis functions. Especially at the lower boundary, the monomial basis shows an exaggerated behavior. However, we must exercise caution since from $L=6$ onwards, we were not able to calculate stable results for the monomial basis. The MSPE for B-splines (0.04574) and Fourier (0.04808) are similar while the monomial basis (0.24181) shows distinctly worse properties. In contrast to the simulation study, the B-splines basis manages to outperform the Fourier basis. Driving factors for this better performance could be that, first, we do not assume a periodic true coefficient function with the same start- and end value and second, that the Fourier basis enforces identical start and end value on $NIR$. It is worthy to note that the reported MSPE for the B-spline and Fourier basis are close to the errors reported in the simulation study for the setup $f_2,Y_1$, which might be caused by the hypothesized similarities between the true coefficient function and $f_2$, but also by a similar high signal-to-noise ratio of the octane numbers.

		\subsubsection{Functional Principal Component Regression}
		The best performance for all numbers of FPC's was achieved with the Fourier Basis, as for the FPCR in the simulation study. In contrast to the simulation study, no evidence of overfitting was found for the best basis choices in any of the three bases. The MSPE for the optimal specification appears to be strictly decreasing in the number of FPC's. As for the simulation study, the interpretation of the results with respect to the chosen basis is difficult: Referring to the plotted estimates $\hat{\beta}(t)$ in figure  \ref{FPCR2_estimate_appl}, \ref{FPCR3_estimate_appl} and \ref{FPCR4_estimate_appl}, it appears to be the case that the higher $n_{FPC}$ is, the more similar behavior of the estimate $\hat{\beta}(t)$ can be observed. Differing from the simulation study, the number of basis functions is steadily increasing for the monomial basis. The behavior for the B-spline basis is similar to the one reported in the simulation study (basis functions increasing in $n_{FPC}$).

@@

	\subsection{Limitations}

@@

	\paragraph{Collinearity Problems in Basis Expansion Regression}
	As already described in the section on the simulation study, the basis expansion regression was in part limited by the numerical instability of the estimation procedure. This is mainly due to an increase in collinearity of the derived regressor matrix shown in Definition \ref{regressor_mat_1}. This problem is inherent to basis systems whose functions are not pairwise orthogonal, such as the monomial or B-splines bases, but gets more pronounced the more functions we add to the basis system and the higher the correlation between those functions. \\
	The numeric instability of the inversion of this matrix makes the estimates unreliable and therefore can make this approach infeasible for non-orthogonal bases in settings where the characteristics of the data set demand a higher number of basis functions than is feasible due to the properties of the estimation procedure.
	\vspace{-0.2cm}

@@

	\subsection{Possible Extensions}

	\paragraph{Orthogonal Polynomials to Solve Collinearity Problems of Monomial Basis}
	To address the collinearity problems described earlier, one possible idea would be to use a system of pair-wise orthogonal polynomials as a basis instead of the monomial basis. One example is the system of Legendre polynomials which are orthogonal by construction and have the same closed span as the monomials. Due to their orthogonality, the problem of collinearity in the regressor matrix are greatly reduced which could allow for larger numbers of basis functions in the basis expansion regression approach. The first eight Legendre polynomials are shown in Appendix section \ref{Basis_Plots}.
	\vspace{-0.2cm}

>>>>

	Due to the implementation of our simulation, we will limit our paper to the case of $\alpha = 0$. However, a different choice of $\alpha$ could lead to improvements in performance of the Monomial basis.
	As the monomials are not pairwise orthogonal, this basis is prone to collinearity problems, which can result in numerically unstable estimates. This restricts the number of basis functions that can be used in the estimation procedures limiting its ability to capture pronounced local peculiarities. These problems will be addressed in more detail in later parts of this paper. The limitation to a low number of monomials can additionally lead to undesirable behavior away from the point of evaluation (cf. \cite{ramsay_functional_2005}).
	\vspace{-0.2cm}

	\paragraph{Fourier Basis}
	In the same way the Monomial basis is connected to the Taylor series, the Fourier basis corresponds to the Fourier series, which can be used to decompose a periodic function into trigonometric functions. Equation \ref{Fourier_Series} shows an example for a function with a period of 1.
	\begin{equation}\label{Fourier_Series}
		s(x) = \frac{A_0}{2} + \sum_{i = 1}^{\infty} A_i \cos(2\pi i x - \phi_i) = \frac{a_0}{2} + \sum_{i = 1}^{\infty}\left[a_i \cos(2\pi i x) + b_i \sin(2\pi i x)\right]
	\end{equation}

@@

	As the B-spline basis does not have infinitely many elements as the full Monomial or Fourier bases do, it is slightly misleading to speak of truncating the B-spline basis at a truncation parameter $L$ in later parts of this paper. For the sake of keeping the notation reasonably concise, we will still keep this notation. However, by convention, truncating a B-spline basis at truncation parameter $L$ shall mean using a B-spline basis consisting of $L$ functions from this point on.

@@

		where $Z \sim \mathcal{N}(0,1)$. In total, we created four combinations, using the two true coefficient functions and the two error terms. These four combinations are then used with a different number of Monomial basis functions $ \in \{1,2, \dots, 6\}$, cubic B-spline basis-function $\{5,6,...,18\}$ and Fourier functions $\{1,3,...,25\}$ to predict the generated responses using the basis expansion approach and the FPCR approach.
		% For the evaluation, we used the prediction MSE calculated by 10 fold cross-validation.
		To obtain valid out of sample properties for the FPCR, within each of the ten ten-fold cross-validation splits, we first calculate the first $n_{FPC}$ FPC's for $n_{FPC} \in \{2,3,4\}$ of the training set $\mathcal{T}$, which was prior smoothed with the respective basis function specification. The obtained Eigenfunctions $\nu^{m,\mathcal{T}}$ are then used to estimate the scores of the holdout set $\mathcal{H}$, $\hat{\xi}_{i}^{m, \mathcal{H}}$  by the equation
		\begin{equation}
			\begin{split}
				\hat{\xi}_{i}^{m, \mathcal{H},L} & =  \int_{0}^{1} {\color{red} \left(X_{i}^{\mathcal{H}}(s) - \hat{\mu}^{\mathcal{T}}(s)\right)} {\color{blue}\tilde{\nu}^{m, L, \mathcal{T}} }\mathrm{d}s) 
			    \approx \int_{0}^{1} {\color{red}\left(\sum_{j = 1}^{L} a_{i,j}^{\mathcal{H}} \phi_j(s)\right)} {\color{blue}\left(\sum_{k = 1}^{L} d_{k}^{m, \mathcal{T}} \phi_{k}(s)\right)} \mathrm{d}s \\
				 &\approx \sum_{j = 1}^{L} \left[ a_{i,j}^{\mathcal{H}}\sum_{k = 1}^{L}  d_{k}^{m, \mathcal{T}} \int_{0}^{1} \phi_j(s) \psi_{k}(s)\mathrm{d}s\right]
			\end{split}
		\end{equation}

@@

	\paragraph{Monomial Basis}
	Due to the in the number of Monomial basis functions increasing collinearity, simulations were conducted up until the sixth monomial, which already shows signs of numerical instability. For reasons outlined in \ref{Monomial_basis}, they seem suited for $f_1$ where it shows a better performance than B-splines for this assumed true coefficient function. A hypothesis for this could be that $f_1$ is an entire function, which can be well approximated with a power series. In case of $f_2$, this basis shows the weakest performance out of all basis, for which the plots in figure \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2} provide evidence. It seems like $\hat{\beta}(t)$ is not changing in the amount of noise and shows exaggerated behaviour at the boundaries. This weakness is especially pronounced in the MSPE for $f_2,Y_1$. For $f_1$, the simulation selects 5(3) and for $f_2$ 5(5) Monomial basis functions for the high(low) signal-to-noise ratio. 
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	Simulations with B-spline basis functions were possible from 4 to 18, since from 18 onwards the simulations were running into collinearity problems.
	Function $f_1$ requires 5(4) B-spline basis functions for the high(low) signal-to-noise ratio to obtain the best fit for the B-spline basis and shows the worst performance of the three basis. An explanation might be its overripe behavior at the boundaries and the exaggeration of the peculiarities of $f_1$ ( figure \ref{basis_expansion_1_1} and \ref{basis_expansion_1_2}), which is especially pronounced for the lower signal-to-noise ratio. For $f_2$, 11(6) B-spline basis functions are chosen for the high(low) signal-to-noise ratio. The B-spline basis outperforms the Monomial basis in $f_2$, but comes second to the Fourier basis. While the basis seems to recognize the peculiarities in $f_2,Y_1$, it seems to struggle with it for the noisy responses in $f_2,Y_2$ ( figure \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2}). With the low signal-to-noise ratio in the responses, the simulation chooses a smaller number of basis functions, which could be to prevent fitting this scalar noise using $\hat{\beta}(t)$.
	\vspace{-0.2cm}

@@

	\paragraph{Monomial Basis}
	For $f_1$, the cross-validated MSPE is decreasing in the number of FPC's and chosen basis functions for both the high and the low signal-to-noise ratio ($4,5,6$ basis functions for $n_{FPC} \in \{2, 3, 4 \}$). In  $f_2$ we also observe a decreasing MSPE in the number of FPC's, but no clear relationship for the chosen basis functions. The Monomial basis shows the weakest performance for all three bases in each setting for all $n_{FPC}$.
	\vspace{-0.2cm}

@@

	\subsection{Interpretation of Results}
		\subsubsection{Basis Expansion Regression}
		The cross-validated MSPE selects 5 basis functions for the Monomial basis. For B-Splines, the cross-validation selects 10 basis functions and 9 basis functions for the Fourier basis. Comparing the estimates in Appendix \ref{Estimates_Appl}, away from the boundaries, the B-spline and Fourier bases show similar behavior, opposing to the Monomial estimate which might be attributable to the lower number of basis functions. Especially at the lower boundary, the Monomial basis shows an exaggerated behavior. However, we must exercise caution since from $L=6$ onwards, we were not able to calculate stable results for the Monomial basis. The MSPE for B-splines (0.04574) and Fourier (0.04808) are similar while the Monomial basis (0.24181) shows distinctly worse properties. In contrast to the simulation study, the B-splines basis manages to outperform the Fourier basis. Driving factors for this better performance could be that, first, we do not assume a periodic true coefficient function with the same start- and end value and second, that the Fourier basis enforces identical start and end value on $NIR$. It is worthy to note that the reported MSPE for the B-spline and Fourier basis are close to the errors reported in the simulation study for the setup $f_2,Y_1$, which might be caused by the hypothesized similarities between the true coefficient function and $f_2$, but also by a similar high signal-to-noise ratio of the octane numbers.

		\subsubsection{Functional Principal Component Regression}
		The best performance for all numbers of FPC's was achieved with the Fourier Basis, as for the FPCR in the simulation study. In contrast to the simulation study, no evidence of overfitting was found for the best basis choices in any of the three bases. The MSPE for the optimal specification appears to be strictly decreasing in the number of FPC's. As for the simulation study, the interpretation of the results with respect to the chosen basis is difficult: Referring to the plotted estimates $\hat{\beta}(t)$ in figure  \ref{FPCR2_estimate_appl}, \ref{FPCR3_estimate_appl} and \ref{FPCR4_estimate_appl}, it appears to be the case that the higher $n_{FPC}$ is, the more similar behavior of the estimate $\hat{\beta}(t)$ can be observed. Differing from the simulation study, the number of basis functions is steadily increasing for the Monomial basis. The behavior for the B-spline basis is similar to the one reported in the simulation study (basis functions increasing in $n_{FPC}$).

@@

	%\subsection{Limitations}

@@

	\paragraph{Collinearity Problems in Basis Expansion Regression}
	As already described in the section on the simulation study, the basis expansion regression was in part limited by the numerical instability of the estimation procedure. This is mainly due to an increase in collinearity of the derived regressor matrix shown in Definition \ref{regressor_mat_1}. This problem is inherent to basis systems whose functions are not pairwise orthogonal, such as the Monomial or B-splines bases, but gets more pronounced the more functions we add to the basis system and the higher the correlation between those functions. \\
	The numeric instability of the inversion of this matrix makes the estimates unreliable and therefore can make this approach infeasible for non-orthogonal bases in settings where the characteristics of the data set demand a higher number of basis functions than is feasible due to the properties of the estimation procedure.
	\vspace{-0.2cm}

@@

	%\subsection{Possible Extensions}

	\paragraph{Orthogonal Polynomials to Solve Collinearity Problems of Monomial Basis}
	To address the collinearity problems described earlier, one possible idea would be to use a system of pair-wise orthogonal polynomials as a basis instead of the Monomial basis. One example is the system of Legendre polynomials which are orthogonal by construction and have the same closed span as the monomials. Due to their orthogonality, the problem of collinearity in the regressor matrix are greatly reduced which could allow for larger numbers of basis functions in the basis expansion regression approach. The first eight Legendre polynomials are shown in Appendix section \ref{Basis_Plots}.
	\vspace{-0.2cm}
